---
title: "Fig 1a - Binary vs All - Trachea dataset"
author: "Jonathan Algoo"
date: "2023-11-06"
output: html_document
---

Fig 1a - Using marker genes that differentiate two cell classes > marker genes that differentiate all cell classes, just try training normal ML classifiers using consensus approach on 1-vs-1 vs marker genes identified for 1-vs-all

Load trachea seurat obj from Tabula Muris dataset
```{r}
library(readr)
library(rsample)
annotations <- read.csv("../scmappr_paper_datasets/annotations_facs.csv") %>% dplyr::filter(tissue == "Marrow")
marrow_ge <- read.csv("../scmappr_paper_datasets/FACS/Marrow-counts.csv")
rownames(marrow_ge) <- marrow_ge$X
marrow_ge %<>% select(-X)
marrow_cell_ids <- marrow_ge %>% colnames()
annotations <- annotations[match(marrow_ge %>% colnames(),annotations$cell),]
celltype_anno <- annotations$cluster.ids
#remove cells with na anno
na_cells <- which(is.na(celltype_anno))
celltype_anno <- celltype_anno[-na_cells]
annotations <- annotations[-na_cells,]
#subset marrow_ge to only include non-na anno
marrow_ge <- marrow_ge[,-na_cells]
#write bpcells for future
marrow_ge_bpcells <- marrow_ge %>% as("Matrix") %>% as("dgCMatrix") %>% as("IterableMatrix") %>% write_matrix_dir(paste0("../scmappr_paper_datasets/bpcells/marrow_tm"), overwrite = T)
```
Convenience function for reading FACS-formatted TM files to bpcells in one command? 

```{r}
#split 80/20 train/test
#get testing and training set in df form
marrow_ge <- t(marrow_ge) %>% as.data.frame()
marrow_ge$cluster.ids <- annotations$cluster.ids
splits <- initial_split(marrow_ge, prop = .80, strata = cluster.ids, pool = 0)
test.set <- splits %>% testing() 
test.set.labels <- test.set$cluster.ids
train.set.labels <- train.set$cluster.ids
test.set <- test.set %>% select(-cluster.ids)
train.set <- train.set %>% select(-cluster.ids)
```


```{r}
tree <- CreateEqualTree(annotations$cluster.ids)
#only sampled 100 cells per class bc small # cells in many cell classes
pairwise_markers <- FindMarkerGenes(ref_bpcells = marrow_ge, tree = tree, ref_metadata = annotations,metadata_cluster_column = "cluster.ids",n_cells_sampled = 100,metadata_cell_label_column = "cell")


all_class_markers <- marker_features(mat = marrow_ge, groups = annotations$cluster.ids, method = "wilcoxon")

#do same processing on all_class_markers as done in pairwise markers in FindMarkerGenes function
all_class_markers %<>% dplyr::filter(foreground_mean > 1 |background_mean > 1) %>% dplyr::select(-background) %>% dplyr::distinct(feature, .keep_all = TRUE) %>% dplyr::mutate(log2_fc = log2(foreground_mean/background_mean))
#get log2fc, and select the top marker genes with the highest abs value log2fc
#have slightly more than 10* number of marker genes bc some genes literally were not expressed in other cell types, so they had log2fc of infinity. Kept all these genes for the present comparison, to make all marker gene group as robust as possible. 
all_class_markers %<>% mutate(abs_log2_fc = log2(foreground_mean/background_mean) %>% abs()) %>% arrange(abs_log2_fc) %>% slice_max(abs_log2_fc, n = 10, by = foreground) %>% pull(feature)
      
all_class_ge <- marrow_ge[all_class_markers,] %>% as.matrix()



      
```

Repeat with 100, 1000 marker genes per broadclass for the all_class_markers later? 